---
title: "Getting Started with corinet"
output: rmarkdown::html_vignette
author: "Liane Muir"
vignette: >
  %\VignetteIndexEntry{Getting Started with corinet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

*corinet* provides a lightweight pipeline for building and analyzing gene
co-expression networks from bulk RNA-seq data. Starting from a
`SummarizedExperiment` object, the package supports the full workflow
from data preparation through network construction, module detection,
visualization, and export.

Let's see what the complete pipeline looks like...

## Installation

```{r install, eval=FALSE}
# Install from GitHub
devtools::install_github("lmuir16/corinet")
```

## Data Preparation

The package includes `example_se`, a `SummarizedExperiment` containing GTEx
skeletal muscle bulk RNA-seq data for 2000 genes across 200 samples.

```{r load}
library(corinet)

data(example_se)
example_se
```

Before building a network, we select the most variable genes and prepare a
log-normalized expression matrix. Working with the top variable genes keeps
pairwise correlation computation tractable and focuses the network on
biologically informative signal.

```{r dataprep}
# Select top 500 most variable genes
se_top <- top_variable_features(example_se, n = 500)

# Optionally subset to a random sample of donors for faster computation
set.seed(42)
se_sub <- select_random_samples(se_top, n = 200)

# Log2-normalize counts and remap Ensembl IDs to HGNC gene symbols
mat <- prepare_expression_matrix(se_sub)

dim(mat)
mat[1:5, 1:3]
```

## Computing Gene-Gene Correlations

With the expression matrix prepared, we compute pairwise Pearson correlations
across all gene pairs. This produces a symmetric genes x genes matrix where
each entry reflects the co-expression strength between two genes.

```{r correlation}
cor_mat <- pairwise_corr(mat, cor_method = "pearson")

dim(cor_mat)
cor_mat[1:5, 1:5]
```

## Building the Adjacency Matrix

We threshold the correlation matrix into a binary adjacency matrix. Gene pairs
with absolute correlation at or above `cor_threshold` are connected by an edge.
Choosing an appropriate threshold is important — too high produces a sparse
disconnected network, too low produces a dense uninformative one. A threshold
of 0.7 is a common starting point for bulk RNA-seq co-expression networks.

```{r adjacency}
adj_mat <- build_adjacency_mat(cor_mat, cor_threshold = 0.7)

# Proportion of possible edges retained
sum(adj_mat) / (nrow(adj_mat) * (nrow(adj_mat) - 1))
```

## Constructing the Network

The adjacency matrix is passed to `build_network()`, which constructs an
igraph object and computes per-gene degree and betweenness centrality.

- **Degree** — the number of other genes a gene is correlated with above the
  threshold. High-degree genes are candidates for hub genes.
- **Betweenness centrality** — how often a gene lies on the shortest path
  between other gene pairs. High betweenness genes act as bridges between
  co-expression modules.

```{r network}
net <- build_network(adj_mat)

# Preview degree distribution
summary(net$degree)
```

## Detecting Gene Modules

Gene modules are groups of highly interconnected genes that likely share
biological function. We use the Louvain algorithm for community detection,
which optimizes modularity — a measure of how well-separated the modules are.

```{r modules}
gene_mods <- detect_network_modules(net, community_method = "louvain")

# Modularity score — higher values indicate clearer module structure
gene_mods$modularity

# Number of modules detected
length(unique(gene_mods$modules))

# Preview module assignments
head(gene_mods$module_df)
```

## Summarizing the Network

`summarize_network()` returns a data frame of global network statistics,
including total edge count, mean degree, hub gene count, and modularity.

```{r summary}
net_summary <- summarize_network(net, gene_mods$modules, gene_mods$modularity)
```

## Identifying Hub Genes

Hub genes are the most highly connected genes in the network. They often
correspond to master regulators or genes central to the biological processes
captured by the co-expression structure.

```{r hubs}
hub_genes <- get_hub_genes(net, modules = gene_mods$modules, n = 20)
head(hub_genes)
```

## Visualization

### Correlation Heatmap

The correlation heatmap shows pairwise gene-gene correlations with hierarchical
clustering, revealing block structure that corresponds to co-expression modules.

```{r heatmap, fig.width=7, fig.height=6}
plot_corr_map(cor_mat, cor_method = "pearson")
```

### Force-Directed Network Graph

The network plot shows the top hub genes as nodes sized by degree and colored
by module, with the top 20 hub genes labeled by gene symbol.

```{r network-plot, fig.width=8, fig.height=7}
plot_network(net,
             modules        = gene_mods$modules,
             n_top          = 80,
             cor_threshold  = 0.7,
             community_method = "louvain")
```

Note that the number of labeled hub genes can be controlled with `n_label` 
to reduce overlap in dense networks:

```{r n_label_example, eval=FALSE}
# Reduce label density for cleaner visualization
plot_network(net,
             modules          = gene_mods$modules,
             n_top            = 80,
             n_label          = 10,
             cor_threshold    = 0.7,
             community_method = "louvain")
```

## Exporting Results

All results can be exported to TSV files for downstream analysis or reporting.
Six files are written to a temporary directory by default.

```{r export}
exported_files <- net_export(
  cor_mat,
  net,
  net_summary,
  gene_mods$modules,
  gene_mods$module_df,
  cor_threshold = 0.7
)

exported_files
```

The exported files are:

- `gene_correlations.tsv` — all gene pairs above the correlation threshold
- `network_summary.tsv` — global network statistics
- `hub_genes.tsv` — top 20 hub genes ranked by degree
- `module_assignments.tsv` — module membership for all genes
- `node_stats.tsv` — degree, betweenness, and module for every gene
- `module_summary.tsv` — per-module size, mean degree, and top hub gene

## Command Line Interface

For non-interactive use, corinet provides a CLI via Rapp. After installing
the package, install the CLI with:

```r
Rapp::install_pkg_cli_apps("corinet")
```

The CLI accepts a genes x samples counts matrix as TSV or CSV, with gene
identifiers as row names. A small example counts matrix is available at
`tests/testdata/counts.tsv` in the package repository for testing.

All subcommands accept the same core arguments:

```bash
# Full network analysis pipeline — exports 6 TSV files
corinet network --counts counts.tsv --output results/ --n-top 500

# Gene correlation heatmap
corinet heatmap --counts counts.tsv --output results/ --n-top 500

# Force-directed network graph
corinet plot-network --counts counts.tsv --output results/ --n-top 500 --n-plot 80
```

See `corinet --help` and `corinet <subcommand> --help` for full parameter
documentation including correlation method, threshold, and community
detection options.

And that's that!

## Session Info

```{r sessioninfo}
sessionInfo()
```
